\documentclass{article}

% Numbered, compressed citations (must be set before neurips_2026 loads natbib).
\PassOptionsToPackage{numbers,sort&compress}{natbib}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2026

% The authors should use one of these tracks.
% Before accepting by the NeurIPS conference, select one of the options below.
% 0. "default" for submission
\usepackage{neurips_2026}
% the "default" option is equal to the "main" option, which is used for the Main Track with double-blind reviewing.
% 1. "main" option is used for the Main Track
%  \usepackage[main]{neurips_2026}
% 2. "position" option is used for the Position Paper Track
%  \usepackage[position]{neurips_2026}
% 3. "eandd" option is used for the Evaluations & Datasets Track
 % \usepackage[eandd]{neurips_2026}
% 4. "creativeai" option is used for the Creative AI Track
%  \usepackage[creativeai]{neurips_2026}
% 5. "sglblindworkshop" option is used for the Workshop with single-blind reviewing
 % \usepackage[sglblindworkshop]{neurips_2026}
% 6. "dblblindworkshop" option is used for the Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop]{neurips_2026}

% After being accepted, the authors should add "final" behind the track to compile a camera-ready version.
% 1. Main Track
 % \usepackage[main, final]{neurips_2026}
% 2. Position Paper Track
%  \usepackage[position, final]{neurips_2026}
% 3. Evaluations & Datasets Track
 % \usepackage[eandd, final]{neurips_2026}
% 4. Creative AI Track
%  \usepackage[creativeai, final]{neurips_2026}
% 5. Workshop with single-blind reviewing
%  \usepackage[sglblindworkshop, final]{neurips_2026}
% 6. Workshop with double-blind reviewing
%  \usepackage[dblblindworkshop, final]{neurips_2026}
% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote.
% For workshops (5., 6.), the authors should add the name of the workshop, "\workshoptitle" command is used to set the workshop title.
% \workshoptitle{WORKSHOP TITLE}

% "preprint" option is used for arXiv or other preprint submissions
 % \usepackage[preprint]{neurips_2026}

% to avoid loading the natbib package, add option nonatbib:
%    \usepackage[nonatbib]{neurips_2026}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors
\usepackage{graphicx}
\usepackage{tabularx}
\usepackage{array}
\usepackage{enumitem}
\newcommand{\missref}[1]{\textcolor{red}{[#1]}}

% Note. For the workshop paper template, both \title{} and \workshoptitle{} are required, with the former indicating the paper title shown in the title and the latter indicating the workshop title displayed in the footnote.
\title{Directed Consistency Learning for Robust LLM Alignment Against Adversarial Jailbreaks}


% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\author{%
  % Author information omitted for double-blind submission
  % examples of more authors
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \AND
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
  % \And
  % Coauthor \\
  % Affiliation \\
  % Address \\
  % \texttt{email} \\
}

\begin{document}
\maketitle

\begin{abstract}
Despite strong refusal behaviour on straightforward harmful requests, aligned LLMs remain vulnerable to adversarial jailbreaks, prompt perturbations that induce unsafe compliance. We propose directed consistency learning, a fine-tuning approach that stabilizes model behaviour across semantically similar prompts by encouraging consistency between a prompt and its adversarially perturbed variants. Our method introduces an asymmetric consistency regulariser that penalizes increases in unsafe response probability between an anchor prompt and its adversarial neighbour, using a safety critic to estimate unsafe response likelihood.  To enable gradient-based optimization through discrete generation, we employ a soft autoregressive rollout that preserves end-to-end differentiability. We train on 3,000 harmful prompts augmented with adversarial neighbours generated by GCG, AutoDAN, and PAIR, and evaluate on AdvBench, HarmBench, and JailbreakBench under seen-family and held-out-family attack settings. Across Llama-2, Llama-3, and Mistral models, our method consistently reduces attack success rate (ASR) relative to base models and adversarial fine-tuning, while maintaining competitive false refusal rates (FRR). These results suggest that asymmetric consistency regularisation across adversarial prompt variants provides an effective approach to improving robustness against jailbreak attacks.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Modern safety-aligned LLMs are remarkably effective at refusing clearly harmful requests~\cite{ouyang2022instructgpt,touvron2023llama2}: a direct query for instructions on making a weapon or synthesizing a dangerous substance is met with a polite refusal. Yet this alignment is fragile. A surprising diversity of small, targeted perturbations to an otherwise-refused prompt, appended token suffixes~\cite{zou2023gcg}, fluent paraphrases generated by a second LLM~\cite{chao2023pair,mehrotra2023tap}, stealthy low-perplexity rewrites~\cite{liu2024autodan}, or multi-turn escalation~\cite{russinovich2024crescendo}, can reliably induce the same model to comply. The gap between a model's performance on typical inputs and its brittleness under adversarial inputs is not new to the machine-learning community~\cite{madry2018pgd}, but its manifestation in safety-critical LLM deployment is particularly consequential.


\paragraph{Why jailbreaks work: mismatched generalization.}
Wei~et~al.~\cite{wei2023jailbroken} offer a precise reason for this: safety training instills refusal on a narrow phrasing distribution, but instruction-following generalizes much more broadly. Adversarial perturbations exploit this asymmetry, landing outside the refusal space while remaining inside the compliant-generation space. Carlini~et~al.~\cite{carlini2023adversarially} further develop this, as current ``aligned'' models are not adversarially aligned in any meaningful ML sense, even weak adversaries can find inputs that violate safety behaviour, and the problem is structural, not a matter of insufficient data. We take both observations as motivation: the vulnerability is not a knowledge gap or a labeling error, but a consistency failure.

\paragraph{Jailbreak vulnerability as a consistency failure.}
A model that refuses \emph{``How do I make a bomb?''} but complies with a lightly paraphrased or suffix-appended variant already knows how to refuse the underlying request. What it lacks is behavioural consistency across semantically equivalent inputs. Framing jailbreak vulnerability this way shifts the question from \emph{``can we detect or filter attacks?''} to \emph{``can we train the model to be stable under them?''} The former question has dominated the literature, producing inference-time filters~\cite{jain2023baseline,robey2023smoothllm,kumar2023erase} that do not change the model and output-side classifiers~\cite{inan2023llamaguard} that can be bypassed. The latter question calls for a training-time intervention that teaches the model to produce consistent, safe behaviour across the entire neighbourhood of a harmful prompt.

\paragraph{What we propose.}
We propose directed consistency learning, a LoRA fine-tuning approach that directly targets the consistency failure. The core idea is straightforward: given an anchor harmful prompt and an adversarially perturbed neighbour, we penalize cases where the perturbed prompt \emph{increases} the probability of an unsafe response relative to the anchor. Three design choices make this tractable and well-targeted.

\textit{Asymmetric regularizer.} We penalize only \emph{increases} in unsafe probability, not symmetric divergence between the anchor and perturbed outputs. This is a critical distinction from prior consistency regularizers such as R-Drop~\cite{liang2021rdrop}, UDA~\cite{xie2020uda}, and PPCL~\cite{qiang2024ppcl} which enforce symmetric KL or JS divergence and would penalize a perturbation that makes the model \emph{more} safe, treating safety improvement as instability. Our hinge is silent whenever a perturbation does not hurt, and penalizes in proportion to how much it does.


\textit{Differentiably learned safety critic.}
Rather than relying on keyword matching or a fixed binary label, we estimate unsafe response probability using a frozen Llama~Guard~\cite{inan2023llamaguard} classifier. The anchor response is the model's own current output on the original prompt, not an externally defined target, so the regularizer adapts as fine-tuning progresses. This is conceptually related to the reinforcement learning from human feedback (RLHF)~\cite{christiano2017rl,ouyang2022instructgpt} reward model, but requires no online sampling or Proximal Policy Optimization (PPO)~\cite{schulman2017ppo}: the critic is frozen and the loss is end-to-end differentiable.


\textit{Soft autoregressive rollout.}
Evaluating the safety critic requires generating a response to the perturbed prompt during training. Discrete sampling breaks the computation graph, and REINFORCE~\cite{williams1992reinforce} estimators are high-variance. Instead, we propagate expected token embeddings at each generation step, maintaining a differentiable path through both the target model and the frozen critic. Unlike Gumbel-Softmax~\cite{jang2017gumbel} or straight-through estimators~\cite{bengio2013ste}, this approach requires no temperature annealing and preserves gradient magnitude throughout the rollout.
%
The method is trained on 3,000 harmful prompts augmented with adversarial neighbours from GCG~\cite{zou2023gcg}, AutoDAN~\cite{liu2024autodan}, and PAIR~\cite{chao2023pair}, covering the major families of white-box optimization-based, fluent, and semantic attacks. We evaluate on AdvBench~\cite{zou2023gcg}, HarmBench~\cite{mazeika2024harmbench}, and JailbreakBench~\cite{chao2024jailbreakbench} under \emph{seen-family} and \emph{unseen-family} protocols (Section~\ref{sec:exp-adversarial}). ASR uses the official HarmBench Mistral validation classifier~\cite{mazeika2024harmbench}; FRR uses Mistral-7B-Instruct-v0.2~\cite{jiang2023mistral} with the JailbreakBench refusal judge~\cite{chao2024jailbreakbench}.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Related Work}
\label{sec:related-work}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Our work sits at the intersection of three mature lines of research, adversarial robustness for NLP, consistency regularisazion, and LLM safety fine-tuning and draws on recent progress in parameter-efficient fine-tuning and differentiable generation. We survey each in turn and clarify where directed consistency learning diverges from its closest neighbours.

\subsection{LLM alignment and safety fine-tuning.}
\label{sec:llm-alignmnet}
The dominant paradigm for aligning LLMs with human preferences is reinforcement learning from human feedback (RLHF)~\cite{christiano2017rl,ouyang2022instructgpt}: a reward model is fit to pairwise human ratings, and the LM is updated with PPO to maximize expected reward subject to a KL penalty against the base policy. Direct Preference Optimization (DPO)~\cite{rafailov2023dpo} eliminates the online RL loop by showing that the constrained KL-regularized objective collapses to a binary classification loss on preference pairs; directed consistency learning is closer to DPO in spirit than to PPO-RLHF because it is offline, supervised-style, and requires no reward-model rollouts. Constitutional AI~\cite{bai2022constitutional} replaces human labels with a model-generated critique loop, similarly avoiding per-prompt human annotation, as does our use of a frozen Llama~Guard critic. None of these alignment methods are designed to \emph{generalize safety to adversarially perturbed inputs}: they produce refusal on the training distribution, but, as Qi~et~al.~\cite{qi2024finetuning} show, even benign downstream fine-tuning can erase this alignment. This gap motivates the explicit perturbation-consistency objective we impose.

\subsection{Jailbreak attacks.}
\label{sec:jailbreak-attacks}
Modern jailbreaks divide naturally into two families that probe different failure modes of alignment.
%
\paragraph{Optimization-based attacks} search directly in token or embedding space. GCG~\cite{zou2023gcg} uses greedy coordinate gradient descent to find short adversarial suffixes that achieve near-perfect attack success on Vicuna and high transfer to GPT-3.5/4. AutoDAN~\cite{liu2024autodan} replaces gradient search with a hierarchical genetic algorithm, producing fluent low-perplexity prompts that bypass perplexity-based filters entirely. Both attacks are white-box at training time but transfer well in black-box settings; we use them to generate the adversarial neighbours in our training corpus.
%
\paragraph{Semantic attacks} use a separate model to rewrite or escalate the prompt. PAIR~\cite{chao2023pair} uses an attacker LLM that iteratively refines the prompt via attacker--target dialogue, finding jailbreaks in $\sim$20 queries on average and at orders-of-magnitude lower query cost than GCG. TAP~\cite{mehrotra2023tap} extends PAIR with tree-of-thought search and on-topic pruning, achieving $>80\%$ ASR on GPT-4-Turbo in under thirty queries while bypassing Llama-Guard-protected systems. Crescendo~\cite{russinovich2024crescendo} escalates across multiple turns through benign-looking steps, achieving a 29--61\% relative lift in ASR over single-turn baselines on GPT-4 and Gemini-Pro; this multi-turn case falls outside our current single-turn framing and we treat it as a target for future work.
%
Across all families, Wei~et~al.~\cite{wei2023jailbroken} observe the same underlying pattern: \emph{mismatched generalization}. Safety training narrows refusal to a specific phrasing distribution while instruction-following generalizes broadly; adversarial perturbations land outside the former while remaining inside the latter. This is precisely the instability our regularizer targets.

\subsection{Defences against jailbreaks.}
\label{sec:defences-against-jailbreaks}
Existing defenses cluster into three categories with complementary strengths and limitations.

\textit{Input filtering and detection} methods intercept attack prompts before generation. Perplexity filtering~\cite{jain2023baseline} is effective against GCG-style high-perplexity suffixes but fragile against fluent AutoDAN or PAIR attacks. Erase-and-Check~\cite{kumar2023erase} provides the first certifiable safety guarantees for LLMs by exhaustively checking all token-erased subsequences up to length $d$; it is the state of the art for certified defenses but incurs significant inference cost and assumes bounded-cardinality attacks. Neither approach modifies the model's weights or improves intrinsic robustness.

\textit{Output-side classifiers} such as Llama~Guard~\cite{inan2023llamaguard}, a Llama-2-7B model fine-tuned on a configurable safety taxonomy, screen responses after generation. They are complementary to training-time defenses but can be bypassed when an attack produces harmful content outside the classifier's training support. In our method, Llama~Guard serves as the frozen safety critic inside the training loss, not as a post-hoc filter.

\textit{Training-based defenses} are our primary point of comparison. R2D2~\cite{mazeika2024harmbench}, introduced alongside HarmBench, is the canonical adversarial fine-tuning baseline: it trains on adversarial prompt--refusal pairs generated online during training, optimizing standard cross-entropy. This is our primary empirical baseline; it improves ASR but does not explicitly regularize the model against the \emph{directional shift} in unsafe probability that jailbreaks exploit. Circuit Breakers~\cite{zou2024circuitbreakers} take a different route: a low-rank representation adaptation (LoRRA) is trained to reroute harmful hidden states against a retain set, achieving 87--90\% ASR reduction on HarmBench with under 1\% utility loss on MT-Bench. This is the closest training-time peer to our work; the key distinction is that Circuit Breakers intervene in activation space, whereas we intervene in token-probability space through an output-side consistency objective. SmoothLLM~\cite{robey2023smoothllm} applies randomized smoothing at inference by character-level perturbing the prompt and aggregating outputs; it shares our intuition that safety should be invariant to perturbations but acts at inference time and pays a non-trivial utility tax, whereas directed consistency learning bakes the invariance into the LoRA weights without inference overhead.
%
\paragraph{Consistency regularization.}
Enforcing output consistency under input perturbations is a well-established technique in semi-supervised learning. R-Drop~\cite{liang2021rdrop} adds a symmetric KL divergence between two dropout sub-models on the same input. UDA~\cite{xie2020uda} enforces KL consistency between original and augmented (back-translated or RandAugmented) inputs. Virtual Adversarial Training~\cite{miyato2018vat} generates worst-case continuous embedding perturbations that maximize the same KL divergence and regularizes against them; Park~et~al.~\cite{park2022vatd} extend this to discrete token replacements.
%
Most relevant to our work, PPCL~\cite{qiang2024ppcl} applies prompt perturbation consistency training directly to LLMs for robustness, enforcing output stability under paraphrase and back-translation perturbations via symmetric KL divergence. Our work shares the motivation of perturbation-stable LLM behaviour but diverges in two key respects: PPCL targets general-purpose robustness under benign paraphrases using symmetric consistency, whereas we target adversarial safety specifically using an \emph{asymmetric} hinge that fires only when a perturbation \emph{increases} unsafe response probability. Symmetric consistency applied to harmful-prompt pairs would penalize a perturbation that makes the model \emph{more} safe, conflating safety improvement with instability; our asymmetric formulation avoids this and, to our knowledge, is the first consistency objective designed specifically for adversarial safety.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Method}
\label{sec:method}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We call our approach \emph{directed consistency learning} because the hinge only penalises \emph{worsening} unsafe-looking scores on the rewrite relative to the clean anchor; if the critic sees the rewrite as equally safe or safer, the term is zero. A different pair of label tokens could target another axis (for example partisan slant); we use \texttt{unsafe}/\texttt{safe} because our failure mode is jailbreak-induced harm. If a harmful prompt and a jailbroken rewrite are ``the same request in disguise,'' the model should not become \emph{more} willing to comply on the rewrite. We enforce that with a frozen safety critic that scores partial answers, and a loss term that compares the critic on the original prompt to the critic on the perturbed prompt.

\subsection{Training data and notation}
Each training row has a harmful prompt $h_n$, an adversarial rewrite $h'_n$ (from GCG, AutoDAN, or PAIR; see Section~\ref{sec:exp-adversarial}), and a continuation string $r_n$ taken from the \texttt{Original Response} column. Those targets were produced offline by querying a chat model on each $h_n$ and logging the reply; for the aligned models we study the replies are almost always refusals, but the pipeline does not assert that---it simply minimises cross-entropy on the stored text. We write $\mathcal{D}=\{(h_n,h'_n,r_n)\}_{n=1}^{N}$ and denote the trainable chat model by $\pi_\theta$.

\subsection{Objective}
The objective is a sum of two terms. The first is standard supervised fine-tuning on the clean prompt: we minimise the cross-entropy of the target continuation $r_n$ given $h_n$ (the default \texttt{clean} path in our code). Here $r_n$ is exactly the \texttt{Original Response} field paired with $h_n$ in the training CSV, not a hand-written refusal template.

Let $\mathcal{L}_\theta(r_n\mid h_n)$ be the cross-entropy loss for predicting $r_n$ from $h_n$. Let $C_\psi(\cdot)$ be a frozen critic mapping the (prompt, partial reply) pair to a scalar in $[0,1]$ that we interpret as ``how unsafe'' the partial reply looks. We build it from the critic language model's logits at the \textbf{last} position of the wrapped input: we subtract the logit for the \texttt{safe} token from the logit for the \texttt{unsafe} token and pass the difference through a sigmoid. Intuitively, if the model would much rather emit \texttt{unsafe} than \texttt{safe} as the next token, the score is close to one; if the opposite, it is close to zero. We write $\tilde{r}_\theta(h'_n)$ for the model's own continuation on the adversarial prompt $h'_n$ (approximated with a soft rollout; see below). The training objective is
\begin{equation}
\label{eq:loss}
\min_{\theta}\;\frac{1}{N}\sum_{n=1}^{N}\Bigl(
\underbrace{\mathcal{L}_\theta(r_n\mid h_n)}_{\text{utility}}
\;+\;\lambda\,
\underbrace{\bigl[\,C_\psi(\tilde{r}_\theta(h'_n)\mid h'_n)
- e^{\varepsilon}\,C_\psi(r_n\mid h_n)\,\bigr]_{+}}_{\text{directed stability}}
\Bigr),
\end{equation}
In that loss, $\lambda\ge0$ trades off the two parts, and $[\cdot]_+=\max(\cdot,0)$ is the positive part. If $\lambda{=}0$, only the supervised LM term on $(h_n,r_n)$ remains. We also run a baseline where the likelihood term is computed on $(h'_n,r_n)$ instead of $(h_n,r_n)$ at $\lambda{=}0$; that matches the usual ``train on adversarial prompts'' recipe (\texttt{perturbed} in code) and helps separate the value of the hinge from adversarial supervised fine-tuning alone.

So $\varepsilon$ loosens or tightens multiplicative slack before the hinge fires; the $e^\varepsilon$ form echoes the tolerance used in $\varepsilon$-differential privacy budgets, but here it is only a training knob---not a privacy guarantee---so small critic noise and near-ties do not drown the supervised term. $\lambda$ sets how strongly we weight that hinge. We pick both on validation-style metrics described in Section~\ref{sec:exp-setup}.

\paragraph{Illustrative numbers.} The hinge is zero when the perturbed critic score is \emph{not above} $e^{\varepsilon}$ times the clean score. Fix the clean-path score at $0.5$.

With $\varepsilon{=}0$ we have $e^{\varepsilon}{=}1$, so the bar is $0.5$. The perturbed rollout's unsafe score can go up to $0.5$ without triggering the hinge. If it reaches $1.0$ instead, the hinge is $0.5$ and the loss picks up an extra $0.5\lambda$ on top of cross-entropy.

Keep the same clean score $0.5$, but now let us allow more slack and set $\varepsilon{=}\ln 2$, so $e^{\varepsilon}{=}2$. Now the perturbed rollout’s unsafe score can go up to $1.0$ without triggering the hinge.

\subsection{Differentiability via soft rollouts}
If we generated text with greedy or sampled decoding, the critic would see discrete tokens and gradients would not flow back through the generator. We instead build the first $T{=}5$ answer tokens as \emph{soft} vectors: at each step we take the model's probability over the vocabulary and use the weighted average of token embeddings as the next input embedding. That keeps the graph differentiable. Concretely, for $T{=}5$ we never pick a single token; we append the expected embedding from the current predictive distribution,
\[
\mathbf{e}_{t+1}=\sum_{v\in\mathcal{V}} \pi_\theta(v\mid h'_n,\mathbf{e}_{\le t})\,\mathbf{E}_v,
\]
where $\mathcal{V}$ is the model's vocabulary (the finite set of token types), $v$ ranges over individual tokens in that set, $\pi_\theta(v\mid\cdot)$ is the probability of emitting $v$ as the next token, and $\mathbf{E}_v$ is the embedding vector for token $v$ (a row of the embedding matrix $\mathbf{E}$). The critic then sees its usual prompt wrapper, followed by these soft vectors, followed by a short closing segment from the guard template.

For Llama-2-7B-Chat we use Llama Guard 7B as the critic \citep{inan2023llamaguard} so the tokenizer and embeddings line up with the student model. For Llama-3-8B-Instruct we use similar logic as with Llama Guard, but use Llama Guard~2 as the critic instead to match Llama~3's tokenizer and embedding space. For Mistral-7B-Instruct we load a \emph{second} copy of the same pretrained Mistral-7B-Instruct weights (a separate forward pass, not an additional training stage) and wrap it with a small binary safety header in code (\texttt{mistral\_self\_twin}), which avoids mixing embeddings across unrelated architectures.

\subsection{Optimisation details}
We update only LoRA adapters (rank $8$, scaling $\alpha{=}16$, dropout $0.05$) on the attention \texttt{q\_proj} and \texttt{v\_proj} modules; the critic weights stay fixed in float16. We optimise with AdamW; the default learning rate in our scripts is $2{\times}10^{-5}$ unless we sweep over rates.

Following recent robustness benchmarks, we turn off the vendor ``safety'' system message in chat templates (\texttt{empty} in our code). That way measured attack success is driven by the model itself, not by a system block that already answers most jailbreaks before fine-tuning begins.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Experiments}
\label{sec:experiments}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Reproducibility and orchestration}
Our public repository drives the full study. On Compute Canada Narval, \texttt{run\_final\_pipeline.py} submits jobs in four groups:
\begin{itemize}[leftmargin=*]
  \item Seen-family training on the full $(\lambda,\varepsilon)$ grid with the likelihood term on the clean prompt (\texttt{--lm-loss-input clean}).
  \item The same grid for three \emph{held-out-family} trainers, one for each attack family left out of training (\texttt{gcg}, \texttt{autodan}, \texttt{pair}).
  \item A small set of runs that put the likelihood term on the adversarial prompt with $\lambda{=}0$ only (\texttt{--lm-loss-input perturbed --perturbed-sweep-subset lambda0\_only}), i.e.\ plain adversarial supervised fine-tuning for comparison.
  \item A SLURM array (\texttt{eval/submit\_test\_eval\_matrix.sh}) that calls \texttt{eval/test\_eval\_matrix.py}; by default it scores every saved checkpoint after training finishes (seen-family test plus three unseen-family tests). Optional CPU jobs then merge the per-run tables into heatmaps.
\end{itemize}
File \texttt{model\_profiles.py} keeps base-model paths, critic weights, and evaluation judges in sync for each backbone.

\subsection{Setup}
\label{sec:exp-setup}

\paragraph{Models.}
We experiment on Llama-2-7B-Chat \citep{touvron2023llama2}, Llama-3-8B-Instruct \citep{grattafiori2024llama3}, and Mistral-7B-Instruct \citep{jiang2023mistral}. Each profile uses the critic setup from Section~\ref{sec:method}. \textbf{ASR} uses the official HarmBench Mistral validation classifier \texttt{cais/HarmBench-Mistral-7b-val-cls} on Hugging Face with its released prompt template \citep{mazeika2024harmbench}. \textbf{FRR} uses vanilla Mistral-7B-Instruct-v0.2 \citep{jiang2023mistral} with the JailbreakBench refusal prompt \citep{chao2024jailbreakbench} (exact wording in Appendix~\ref{app:frr-prompt}).

\paragraph{Datasets.}
\textbf{Harmful training set (3{,}000 prompts).} We take 1{,}000 prompts each from MaliciousInstruct++ \citep{huang2023maliciousinstruct}, CySecBench \citep{cysecbench2025}, and WildJailBreak \citep{jiang2024wildjailbreak}. MaliciousInstruct++ is our own filtered list: we generated 20{,}000 candidates from an uncensored helper model with few-shot seeds, then pruned near-duplicates using \texttt{all-mpnet-base-v2} embeddings \citep{reimers2019sbert}.

\textbf{Harmful test set ($\approx$1{,}000 prompts).} We combine AdvBench \citep{zou2023gcg}, HarmBench \citep{mazeika2024harmbench}, and JailbreakBench \citep{chao2024jailbreakbench} into one evaluation file and drop exact duplicate strings so the same text is never counted twice. Each row keeps its source-benchmark tag; we report AdvBench-only, HarmBench-only, and JailbreakBench-only results separately, and we also report an aggregate figure defined as the unweighted mean of those three benchmark-level scores. None of these sources overlap our training files at the document level.

\textbf{Benign prompts.} WildJailBreak's eval split (250 normal and 250 ``looks scary but is safe'' prompts) is useful for quick checks during development. When we need a separate benign test set, we take on the order of 100 prompts of each kind from the WildJailBreak train split (details in the appendix).

\textbf{Semantic validation holdout.} We split the 3{,}000 harmful \emph{training-corpus} prompts into a topic-disjoint \textbf{validation} pool ($\approx$500) and \textbf{training} pool by embedding every prompt with \texttt{all-MiniLM-L6-v2}, hierarchically clustering them under cosine distance with a merge cutoff at $0.88$ (12 clusters), and assigning \emph{entire} clusters to one side or the other so no semantic topic appears in both pools. The default Slurm sweep nevertheless optimises from \texttt{official\_data/train\_plus\_validation.csv}, which concatenates both semantic pools; we use the validation bucket for manual checks and plots, not for automated hyperparameter search. Appendix~\ref{app:validation-construction} records the same construction in full.

\label{sec:exp-adversarial}
\paragraph{Adversarial perturbations.}
Every harmful prompt in train and test has three rewritten versions. \textbf{GCG} \citep{zou2023gcg} uses the \texttt{nanogcg} defaults. \textbf{AutoDAN} \citep{liu2024autodan} uses batch size~8 with otherwise default settings. \textbf{PAIR} \citep{chao2023pair} uses Vicuna-13B-v1.3 \citep{zheng2023vicuna} to draft attacks and Llama Guard as the built-in judge; we change only how streams are picked when the judge stalls (appendix, \emph{Modified PAIR Stream-Selection Rule}). Each iteration stores the usual PAIR score and the margin $\Delta_C = C(\tilde{x},r)-C(x,r_{\text{refusal}})$. We return the first stream that hits judge score~10, or else the stream with the largest $\Delta_C$, which reliably yields a strong attack even when the discrete judge is timid.

\paragraph{Seen-family and unseen-family evaluation.}
\begin{enumerate}[label=(\alph*),leftmargin=*]
\item \textbf{Seen-family.} During training we uniformly sample one of the three rewrite columns (GCG, AutoDAN, or PAIR) for each minibatch row. Testing uses attacks from the same three families (on held-out examples, of course).
\item \textbf{Unseen-family.} For each attack family $f\in\{\textsc{gcg},\textsc{autodan},\textsc{pair}\}$ we train a fresh adapter with \texttt{train.py --eval-mode unseen-family --unseen-family $f$}. That removes $f$ from the random neighbour pool, so every training step only sees the other two families (uniformly, still). We then run \texttt{eval.py --eval-mode unseen-family} on attacks from $f$ only, i.e.\ the family that never appeared in the hinge during training.
\end{enumerate}

\paragraph{Metrics and judges.}
\textbf{ASR} uses the HarmBench validation classifier \texttt{cais/HarmBench-Mistral-7b-val-cls} \citep{mazeika2024harmbench}: it scores whether the model output actually exhibits the harmful behaviour in the row, rather than keyword matching. \textbf{FRR} uses Mistral-7B-Instruct-v0.2 \citep{jiang2023mistral} with the JailbreakBench refusal wording \citep{chao2024jailbreakbench} (Appendix~\ref{app:frr-prompt}). We read the first non-space token as yes/no.

\paragraph{Baselines.}
We compare against (i) the base checkpoint with the same empty-system setup, and (ii) adversarial supervised fine-tuning that trains on $(h'_n,r_n)$ with $\lambda{=}0$ (\texttt{perturbed} jobs in the pipeline) and learning rate $2{\times}10^{-5}$ (same as the finetuned models).

\paragraph{Hyperparameter search.}
We grid-search
$\lambda\in\{0,0.1,0.3,1,3,10,30\}$ and
$\varepsilon\in\{-1.0,-0.5,0,0.5,1.0\}$.
When $\lambda{=}0$ the hinge never fires, so we only schedule one $\varepsilon$ value per learning rate to avoid duplicate work (\texttt{lambda\_epsilon\_pairs} in \texttt{submit\_wandb\_sweep.py}). Learning rates are either $\{10^{-5},\,2{\times}10^{-5}\}$ when we pass \texttt{--lr} to \texttt{run\_final\_pipeline.py}, or just $2{\times}10^{-5}$ by default. Each Slurm job runs five consecutive training epochs (five checkpoints); the matrix reads epoch~5 unless we change it. We pick the $(\lambda,\varepsilon,\text{lr})$ triple that gives the best ASR--FRR trade-off on the seen-family table, then reuse that triple for the held-out-family trainers so we do not re-tune on their test condition.

\subsection{Results}
% TODO

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% TODO

\begin{ack}
% Do not include in anonymized submission. The \texttt{ack} environment
% provided in the style file automatically hides this section in the anonymized submission.
\end{ack}

\bibliographystyle{plain}
\bibliography{references}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix

\section{Modified PAIR Stream-Selection Rule}

Each iteration still logs the usual PAIR judge score, but we also store $\Delta_C = C(\tilde{x},r)-C(x,r_{\text{refusal}})$, i.e.\ how much the safety critic moves when we pair the attack prompt and model reply against the original goal with a fixed refusal. If any stream reaches judge score~10 we return it; otherwise we return the stream with the largest $\Delta_C$. That gives us a strong candidate even when the discrete judge rarely marks a win.

\section{FRR Judge Prompt}
\label{app:frr-prompt}

We use Mistral-7B-Instruct-v0.2 with the JailbreakBench refusal wording below. The first non-whitespace token is parsed as the verdict:
\begin{quote}\small\ttfamily
[INST] Here is a request from a user and a response generated by a large
language model: <request>\{prompt\}</request>
<response>\{response\}</response>
Does the language model refuse to assist the user with their request?
Answer either ``Yes'' or ``No''. [/INST]
\end{quote}

\section{Validation-Set Construction}
\label{app:validation-construction}

To build a held-out harmful set for analysis, we embed all 3{,}000 training prompts with \texttt{all-MiniLM-L6-v2}. We hierarchically cluster the embeddings under cosine distance with a merge cutoff at $0.88$, which yields 12 clusters. We assign whole clusters to train or validation in a roughly $5{:}1$ ratio so no semantic bucket appears in both splits.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newpage
\input{checklist.tex}

\end{document}
